---
title: "Carbon boxes"
output: 
  html_document:
    df_print: paged
    code_folding: hide
author: "John Favate"
date: "`r Sys.time()`"
---

<style type="text/css">
.main-container {
  max-width: 1500px;
  margin-left: auto; 
  margin-right: auto;
}
</style>

```{r}
# Prevent printing of warnings and such in the HTML
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.align = "center")
```

```{r}
library(tidyverse)
library(ggpubr)
library(patchwork)
library(scales)
```

Load some functions required for making graphs for this figure
```{r}
source("./figure_funcs.R")
```

```{r}
df <- read_csv("../../data_frames/targeted_with_imps.csv") %>% 
  filter(is_standard == FALSE) %>% 
  unite("cc", compound, charge, sep = " ", remove = FALSE) %>% 
  unite("ccp", compound, charge, phase, sep = " ", remove = FALSE)

mut.lines <- c("A-1", "A-2", "A-3", "A-4", "A+3", "A+6")

clean.names <- data.frame(line = unique(df$line)) %>%
  mutate(clean = case_when(
    grepl("ap", line) ~ str_replace(line, "ap", "A+"),
    grepl("am", line) ~ str_replace(line, "am", "A-"),
    grepl("rel", line) ~ str_replace(line, "rel60", "R0")
  ))

# combine them and change the one compounds name so it matches the name on fig1
df <- left_join(df, clean.names) %>% 
  mutate(is_mutator = clean %in% mut.lines,
         compound = ifelse(compound == "Nicotinamide ribotide", "Nicotinamide\nmononucleotide", compound))
```

### dotplots

A function for graphing. This function also exists in the function file that was loaded, but it needed to be modified to produce a guide for this specific figure, the rest of them may not need that.
```{r}
dotplot_graph <- function(x,n) {
  # get the comps and clean it up
  comp.df <- df %>% 
    filter(compound %in% x) %>% 
    mutate(charge = ifelse(charge == "positive", "(+)", "(-)"),
           phase = ifelse(phase == "e", "Exponential", "Stationary"))
  
  # get evo values
  evo.pa <- comp.df %>%
    filter(!grepl("R0", clean)) %>%
    group_by(clean, phase, compound, charge) %>%
    summarise(evompa = mean(n_peak_area)) %>%
    ungroup()
  
  # get ancestral values
  anc.pa <- comp.df %>%
    filter(grepl("R0", clean)) %>%
    group_by(phase, compound, charge) %>%
    summarise(ancmpa = mean(n_peak_area)) %>%
    ungroup()
  
  # combine and calc changes
  fc.df <-
    left_join(evo.pa, anc.pa, by = c("phase", "compound", "charge")) %>%
    mutate(l2rat = log2(evompa / ancmpa),
           compound = factor(compound, rev(x)))
  
  # plot it
  p <- fc.df %>%
    ggplot(., aes(l2rat, compound, color = charge)) +
    geom_vline(aes(xintercept = 0), linetype = 5)+
    geom_boxplot(outlier.shape = NA)+
    geom_point(position = position_dodge(width = .9), alpha = .25) +
    facet_wrap( ~ phase, ncol = n, scales = "free_x") +
    theme_bw() +
    theme(
      text = element_text(size = 12),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    ) +
    labs(x = expression(paste(log[2], "(Evo peak area / Anc peak area)")),
         y = NULL)+
    scale_color_manual(values = c("black", "red"), name = "Ionization mode")
  
  return(p)
}
```


This is a named list of compounds I want in the plot, the name is arbitrary but describes the pathway
```{r}
clist <- list(
  "carbon" = c(
    "Glucose",
    "Glucose-6-phosphate",
    "Fructose-6-phosphate",
    "Fructose-1 6-bisphosphate",
    "Dihydroxyacetone phosphate",
    "D-Glyceraldehyde-3-phosphate",
    "Glycerol-3-phosphate",
    "3-Phosphoglycerate",
    "Phosphoenolpyruvate",
    "Pyruvate",
    "Aconitate",
    "Isocitrate",
    "a-ketoglutarate",
    "Succinyl-CoA",
    "Succinate",
    "Fumarate",
    "Malate",
    "Acetyl CoA"
  )
)
```

```{r}
dp.list <- lapply(clist, dotplot_graph, n=2)

dp.list
```

```{r}
ff <- dp.list[[1]]

ff
```

```{r}
range_func(c(
    "Glucose",
    "Glucose-6-phosphate",
    "Fructose-6-phosphate",
    "Fructose-1 6-bisphosphate",
    "Dihydroxyacetone phosphate",
    "D-Glyceraldehyde-3-phosphate",
    "Glycerol-3-phosphate",
    "3-Phosphoglycerate",
    "Phosphoenolpyruvate",
    "Pyruvate",
    "Aconitate",
    "Isocitrate",
    "a-ketoglutarate",
    "Succinyl-CoA",
    "Succinate",
    "Fumarate",
    "Malate",
    "Acetyl CoA"
  ))
```


```{r}
ggsave(plot = ff, filename = "../../figures/carbon_boxes.pdf", width = 6.51, height = 6)
```


