---
title: "Data comparisons"
output: 
  html_document:
    df_print: paged
    code_folding: hide
author: "John Favate"
date: "`r Sys.time()`"
---

<style type="text/css">
.main-container {
  max-width: 1500px;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r}
# Prevent printing of warnings and such in the HTML
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.align = "center")
```

```{r}
library(tidyverse)
library(patchwork)
library(scales)
library(parallel)
library(broom)
library(ggpubr)
library(corrr)
library(ggrepel)

df <- read_csv("../../data_frames/targeted_with_imps.csv") %>% 
  filter(is_standard == FALSE)

mut.lines <- c("A-1", "A-2", "A-3", "A-4", "A+3", "A+6")

clean.names <- data.frame(line = unique(df$line)) %>%
  mutate(clean = case_when(
    grepl("ap", line) ~ str_replace(line, "ap", "A+"),
    grepl("am", line) ~ str_replace(line, "am", "A-"),
    grepl("rel", line) ~ str_replace(line, "rel60", "R0")
  ))

df <- left_join(df, clean.names) %>% 
  mutate(is_mutator = clean %in% mut.lines)
```

### Growth phase specific PCA

Shape the input data as a list, one per growth phase
```{r}
pca.input.list <- df %>%
  select(-is_standard, -was_imputed, -peak_area, -is_mutator) %>%
  mutate(n_peak_area = log10(n_peak_area)) %>%
  unite("cc", compound, charge, sep = "_") %>%
  group_by(line, phase, cc) %>% 
  summarise(mpa = mean(n_peak_area)) %>% 
  ungroup() %>% 
  mutate(phase = ifelse(phase == "e", "Exponential", "Stationary")) %>% 
  split(.$phase) %>% 
  map(function(x){
    x %>% 
      select(-phase) %>% 
      pivot_wider(names_from = cc, values_from = mpa) %>% 
      left_join(clean.names) %>% 
      column_to_rownames("clean") %>% 
      select(-line)
  })
```

Run the PCAs
```{r}
pca.res.list <- lapply(pca.input.list, prcomp, scale = TRUE)
```

Code to make the graphs
```{r}
p.list <-
  lapply(setNames(names(pca.res.list), names(pca.res.list)), function(x) {
    # the results
    res <- pca.res.list[[x]]
    
    # the input
    inp <- pca.input.list[[x]]
    
    # var explained
    var.explained <- res %>%
      tidy(matrix = "eigenvalues")
    
    # create new axis labels
    pc1.explain <-
      paste0("PC1 (", 100 * signif(var.explained[var.explained$PC == 1, ]$percent, 4), "%)")
    
    pc2.explain <-
      paste0("PC2 (", 100 * signif(var.explained[var.explained$PC == 2, ]$percent, 4), "%)")
    
    # the plot df
    pdf <- res %>%
      augment(inp) %>%
      select(`.rownames`, starts_with(".fitted")) %>%
      rename("line" = ".rownames") %>%
      mutate(phase = x,
             is_mutator = line %in% mut.lines)
    
    # color
    if (x == "Exponential") {
      colorz <- "royalblue"
    } else {
      colorz <- "goldenrod"
    }
    
    # the plot
    ggplot(pdf, aes(`.fittedPC1`,
                    `.fittedPC2`,
                    label = line,
                    color = is_mutator)) +
      geom_point(size = 1)+
      geom_text_repel(
        seed = 1,
        min.segment.length = 1e-12,
        segment.alpha = .5,
        max.overlaps = Inf,
        show.legend = FALSE,
        size = 2.5,
      ) +
      theme_bw() +
      theme(
        text = element_text(size = 12),
        panel.grid = element_blank()
      ) +
      labs(x = pc1.explain,
           y = pc2.explain) +
      facet_wrap( ~ phase)+
      scale_color_manual(values = c("steelblue4", "orange4"),
                         name = NULL,
                         labels = c("Non-mutator",
                                    "Mutator"))
  })
```

```{r fig.width = 5, fig.height = 2.75}
pcas <- (p.list[[1]] | p.list[[2]]) + plot_layout(guides = "collect") & theme(legend.position = "bottom",
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(-10,-10,-10,-10))

pcas
```

Save these results for use in another figure
```{r}
write_rds(pca.input.list, "../../rds_files/fig4_pca_input_list.rds")
write_rds(pca.res.list, "../../rds_files/fig4_pca_res_list.rds")
```

### Extent of parallelism in fold-changes to the ancestor

The code for this is in a separate Rscript that calculates both the actual and randomized set of correlations. That Rscript is at `../analysis/randomizations_for_fig2.R`. Run that separately and it will deposit a file `../../data_frames/randomizations_for_fig2.csv.gz` which is loaded here. It's quite large so it takes some time. This may not work right out of the box if your OS doesn't handle gzip compression well.

Load the randomizations
```{r}
combo.cors <- read_csv("../../data_frames/randomizations_for_fig2.csv.gz")
```

KS tests to test for differences
```{r}
ks.res <- combo.cors %>%
  split(.$phase) %>%
  map(function(x) {
    obs <- x[x$iter == "Observed", ]$R
    
    exp <- x[x$iter == "Expected", ]$R
    
    ks.test(obs, exp) %>%
      tidy() %>%
      mutate(cuts = cut(
        p.value,
        c(1, .05, .01, .001, .0001,-Inf),
        labels = c("****", "***", "**", "*", "NS")
      ))
  }) %>% 
  bind_rows(.id = "phase")

ks.res
```

Also test the difference between Obs Ex and Obs St
```{r}
ks.test(
  combo.cors[combo.cors$phase == "Ex" & combo.cors$iter == "Observed",]$R,
  combo.cors[combo.cors$phase == "St" & combo.cors$iter == "Observed",]$R
  )
```

```{r}
range(combo.cors$points, na.rm = TRUE)
```

This takes really long to draw because there's so many points in the randomizations boxes.
```{r fig.width = 2, fig.height = 3}
p.cors <- combo.cors %>%
  ggplot(.) +
  geom_jitter(aes(phase, points, color = iter), position=position_jitterdodge(jitter.height = 0), size = .2)+
  geom_boxplot(fill = NA, aes(phase, R, color = iter), outlier.size = .2)+
  theme_bw()+
  theme(text = element_text(size = 12),
        panel.grid = element_blank())+
  labs(x = NULL,
       y = expression(paste(rho, " of ", log[2], "(FC)", sep = "")))+
  scale_color_manual(values = c("grey60", "black"), name = NULL, guide = "none")+
  geom_text(inherit.aes = FALSE, data = ks.res, aes(phase, .88, label = cuts))+
  annotate("text", x = 2, y = 1.05, label = "Observed", size = 2.5)+
  annotate("text", x = 1, y = 1.05, label = "Expected", color = "grey60", size = 2.5)
```

### Cor network plot

Need fold changes
```{r}
# these are the ancestral mean peak areas
anc.vals <- df %>% 
  mutate(age = ifelse(grepl("R0", clean), "Ancestor", "Evolved")) %>% 
  filter(age == "Ancestor") %>% 
  group_by(charge, phase, compound) %>% 
  summarise(anc_mean = mean(n_peak_area)) %>% 
  ungroup()

# these are the evo mean peak areas
evo.vals <- df %>% 
  mutate(age = ifelse(grepl("R0", clean), "Ancestor", "Evolved")) %>% 
  filter(age != "Ancestor") %>% 
  group_by(charge, phase, compound, clean) %>% 
  summarise(evo_mean = mean(n_peak_area)) %>% 
  ungroup()

# combine them and calculate fold changes
fc.df <- left_join(evo.vals, anc.vals) %>% 
  mutate(l2ratio = log2(evo_mean / anc_mean)) %>% 
  select(charge, phase, compound, clean, l2ratio)
```


Instead, one of these cor network plots.
```{r fig.width = 4, fig.height = 4}
cor.net.list <- fc.df %>%
  split(.$phase) %>%
  map(function(x) {
    x %>%
      select(-phase) %>%
      unite("comp", charge, compound, sep = "_") %>%
      pivot_wider(names_from = clean, values_from = l2ratio) %>%
      column_to_rownames("comp") %>%
      correlate(method = "spearman") %>%
      network_plot(
        min_cor = 0,
        colors = viridis::plasma(64),
        legend = "range",
        repel = FALSE
      )
  })
  
cor.net.list
```

### Final figure

```{r fig.width = 6.51, fig.height = 5}
top <- (pcas | p.cors) + plot_layout(widths = c(.35, .35, .3))

bottom <- cor.net.list$e | cor.net.list$s

ff <- (top / bottom) + plot_annotation(tag_levels = "A") + plot_layout(heights = c(.35, .65))

ff
```

Save it
```{r}
ggsave(plot = ff, filename = "../../figures/data_comparisons.pdf", width = 6.51, height = 5)
```