---
title: "NAD supplement"
output: 
  html_document:
    df_print: paged
    code_folding: hide
author: "John Favate"
date: "`r Sys.time()`"
---

<style type="text/css">
.main-container {
  max-width: 1500px;
  margin-left: auto; 
  margin-right: auto;
}
</style>

```{r}
# Prevent printing of warnings and such in the HTML
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.align = "center")
```

```{r}
library(tidyverse)
library(ggpubr)
library(patchwork)
library(scales)
```

Load some functions required for making graphs for this figure
```{r}
source("./figure_funcs.R")
```

```{r}
df <- read_csv("../../data_frames/targeted_with_imps.csv") %>% 
  filter(is_standard == FALSE) %>% 
  unite("cc", compound, charge, sep = " ", remove = FALSE) %>% 
  unite("ccp", compound, charge, phase, sep = " ", remove = FALSE)

mut.lines <- c("A-1", "A-2", "A-3", "A-4", "A+3", "A+6")

clean.names <- data.frame(line = unique(df$line)) %>%
  mutate(clean = case_when(
    grepl("ap", line) ~ str_replace(line, "ap", "A+"),
    grepl("am", line) ~ str_replace(line, "am", "A-"),
    grepl("rel", line) ~ str_replace(line, "rel60", "R0")
  ))

# combine them and change the one compounds name so it matches the name on fig1
df <- left_join(df, clean.names) %>% 
  mutate(is_mutator = clean %in% mut.lines,
         compound = ifelse(compound == "Nicotinamide ribotide", "Nicotinamide\nmononucleotide", compound))
```

### dotplots

A function for graphing. This function also exists in the function file that was loaded, but it needed to be modified to produce a guide for this specific figure, the rest of them may not need that.
```{r}
dotplot_graph <- function(x,n) {
  # get the comps and clean it up
  comp.df <- df %>% 
    filter(compound %in% x) %>% 
    mutate(charge = ifelse(charge == "positive", "(+)", "(-)"),
           phase = ifelse(phase == "e", "Exponential", "Stationary"))
  
  # get evo values
  evo.pa <- comp.df %>%
    filter(!grepl("R0", clean)) %>%
    group_by(clean, phase, compound, charge) %>%
    summarise(evompa = mean(n_peak_area)) %>%
    ungroup()
  
  # get ancestral values
  anc.pa <- comp.df %>%
    filter(grepl("R0", clean)) %>%
    group_by(phase, compound, charge) %>%
    summarise(ancmpa = mean(n_peak_area)) %>%
    ungroup()
  
  # combine and calc changes
  fc.df <-
    left_join(evo.pa, anc.pa, by = c("phase", "compound", "charge")) %>%
    mutate(l2rat = log2(evompa / ancmpa),
           compound = factor(compound, rev(x)))
  
  # plot it
  p <- fc.df %>%
    ggplot(., aes(l2rat, compound, color = charge)) +
    geom_vline(aes(xintercept = 0), linetype = 5)+
    geom_point(position = position_dodge(width = .9), alpha = .25) +
    facet_wrap( ~ phase, ncol = n) +
    theme_bw() +
    theme(
      text = element_text(size = 12),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank()
    ) +
    labs(x = expression(paste(log[2], "(Evo peak area / Anc peak area)")),
         y = NULL)+
    scale_color_manual(values = c("black", "red"), name = "Ionization\nmode")
  
  return(p)
}
```


This is a named list of compounds I want in the plot, the name is arbitrary but describes the pathway
```{r}
clist <- list(
  "NAD" = c(
    "Aspartate",
    "NAD+",
    "NADH",
    "NADPH",
    "NADP+",
    "Nicotinamide\nmononucleotide",
    "Nicotinate"
  )
)
```

```{r}
dp.list <- lapply(clist, dotplot_graph, n=2)

dp.list.med <- lapply(clist, dotplot_med_graph)

dp.list
dp.list.med
```

This gives the range of changes over both ionization modes.
```{r}
lapply(clist, range_func)
```

### cor between NAD/NADP

Get fold changes
```{r}
comp.df <- df %>%
  filter(compound %in% clist$NAD) %>%
  mutate(
    charge = ifelse(charge == "positive", "(+)", "(-)"),
    phase = ifelse(phase == "e", "Exponential", "Stationary")
  )

# get evo values
evo.pa <- comp.df %>%
  filter(!grepl("R0", clean)) %>%
  group_by(clean, phase, compound, charge) %>%
  summarise(evompa = mean(n_peak_area)) %>%
  ungroup()

# get ancestral values
anc.pa <- comp.df %>%
  filter(grepl("R0", clean)) %>%
  group_by(phase, compound, charge) %>%
  summarise(ancmpa = mean(n_peak_area)) %>%
  ungroup()

# combine and calc changes
fc.df <-
  left_join(evo.pa, anc.pa, by = c("phase", "compound", "charge")) %>%
  mutate(l2rat = log2(evompa / ancmpa))
```

```{r fig.width = 8, fig.height = 8}
nad.combs <- combn(
  c("NADH", "NAD+", "NADP+", "NADPH"),
  2,
  simplify = FALSE
)

plist <- lapply(nad.combs, function(x) {
  fc.df %>%
    filter(compound %in% x & phase == "Exponential") %>%
    select(clean, charge, compound, l2rat) %>%
    pivot_wider(names_from = compound, values_from = l2rat) %>%
    ggplot(., aes(!!sym(x[1]),!!sym(x[2]), color = charge)) +
    geom_abline(aes(slope = 1, intercept = 0), linetype = 5) +
    geom_point(alpha = .5) +
    stat_cor(size = 3, aes(label = ..r.label..), show.legend = FALSE) +
    theme_bw() +
    theme(
      text = element_text(size = 12),
      panel.grid.major.x = element_blank(),
      panel.grid = element_blank()
    ) +
    scale_color_manual(values = c("black", "red"), name = "Ionization mode", guide = "none")
})

nad.cors <- wrap_plots(plist, ncol = 2)
```

### change across growth phases

```{r}
xlabs <- c(
  "GS Ex",
  "GS St",
  "GSSG Ex",
  "GSSG St",
  "AS Ex",
  "AS St"
)

fac.labs <- c("positive" = "Positive mode", "negative" = "Negative mode")

p.boxes <- df %>% 
  filter(compound %in% c("NAD+", "NADH", "NADP+", "NADPH")) %>% 
  group_by(charge, clean, phase, compound) %>% 
  summarise(mpa = mean((n_peak_area))) %>% 
  ungroup() %>% 
  mutate(phase = ifelse(phase == "e", "Ex", "St"),
         paired = paste(clean, compound),
         colz = ifelse(grepl("R0", clean), "anc", "evo")) %>% 
  arrange(desc(colz)) %>% 
  ggplot(., aes(interaction(phase, compound), mpa))+
  geom_boxplot(color = "grey70", outlier.size = NULL)+
  geom_point(aes(color = colz), size = .5)+
  geom_line(aes(group = interaction(paired, compound), color = colz), alpha = .5)+
  facet_wrap(~charge, labeller = as_labeller(fac.labs), ncol = 1)+
  theme_bw()+
  theme(panel.grid = element_blank(),
        text = element_text(size = 13),
        legend.position = c(.8, .93),
        legend.background = element_blank(),
        axis.text.x = element_text(angle = 30, hjust = 1))+
  scale_x_discrete()+
  labs(x = NULL,
       y = "Normalized peak area")+
  scale_color_manual(values = c("chocolate1", "cadetblue"), name = NULL, labels = c("Ancestor", "Evolved"))+
  scale_y_log10(labels = trans_format("log10", math_format(10^.x)))

p.boxes
```

### Final figure

This is a pretty hacky use of various patchwork functions but without this, the long text labels of panel A force the y axis labels for panel B to be extremely far from the graph.
```{r width = 6.5, height = 7}
top <- dp.list$NAD

bottom <- nad.cors | p.boxes

# ff <- (top / wrap_elements(full = bottom)) + plot_layout(heights = c(.3, .7)) + plot_annotation(tag_levels = "A")

ff <- (top / bottom) + plot_layout(heights = c(.3, .7)) + plot_annotation(tag_levels = "A")
```

```{r}
bottom <- cowplot::plot_grid(nad.cors, p.boxes, labels = c("B", "C"), label_fontface = "plain")

ff <- cowplot::plot_grid(dp.list$NAD, bottom, ncol = 1, labels = c("A", "", ""), rel_heights = c(.35, .65), label_fontface = "plain")

ff
```


```{r}
ggsave(ff, filename = "../../figures/nad_supp.pdf", width = 6.51, height = 7)
```

