---
title: "Clustering"
output: 
  html_document:
    df_print: paged
    code_folding: hide
author: "John Favate"
date: "`r Sys.time()`"
---

<style type="text/css">
.main-container {
  max-width: 1500px;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r}
# Prevent printing of warnings and such in the HTML
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.align = "center")
```

Load packages
```{r}
library(tidyverse)
library(sinib)
library(scales)
library(broom)
library(patchwork)
```

Read in the data, converting the names to cleaner forms and marking muators in the process.
```{r}
df <- read_csv("../../data_frames/targeted_with_imps.csv") %>% 
  filter(is_standard == FALSE)

clean.names <- data.frame(line = unique(df$line)) %>%
  mutate(clean = case_when(
    grepl("ap", line) ~ str_replace(line, "ap", "A+"),
    grepl("am", line) ~ str_replace(line, "am", "A-"),
    grepl("rel", line) ~ str_replace(line, "rel60", "R0")
  ))

mut.lines <- c("A-1", "A-2", "A-3", "A-4", "A+3", "A+6")

df <- left_join(df, clean.names)
```

Get fold changes from ancestor to each evolved line
```{r}
# these are the ancestral mean peak areas
anc.vals <- df %>% 
  mutate(age = ifelse(grepl("R0", clean), "Ancestor", "Evolved")) %>% 
  filter(age == "Ancestor") %>% 
  group_by(charge, phase, compound) %>% 
  summarise(anc_mean = mean(n_peak_area)) %>% 
  ungroup()

# these are the evo mean peak areas
evo.vals <- df %>% 
  mutate(age = ifelse(grepl("R0", clean), "Ancestor", "Evolved")) %>% 
  filter(age != "Ancestor") %>% 
  group_by(charge, phase, compound, clean) %>% 
  summarise(evo_mean = mean(n_peak_area)) %>% 
  ungroup()

# combine them and calculate fold changes
fc.df <- left_join(evo.vals, anc.vals) %>% 
  mutate(l2ratio = log2(evo_mean / anc_mean),
         sig = abs(l2ratio) >= 1) %>% 
  select(charge, phase, compound, clean, l2ratio, sig) %>% 
  unite("cc", charge, compound)
  

fc.df
```

fc distributions
```{r}
ggplot(fc.df, aes(l2ratio))+
  geom_histogram(binwidth = .5, boundary = 0)+
  geom_vline(aes(xintercept = -1))+
  geom_vline(aes(xintercept = 1))+
  scale_x_continuous(breaks = seq(-10,10,1))+
  facet_wrap(~phase, ncol = 1)
```

how many metabolite features (charge + compound) are there?
```{r}
total.metabs <- length(unique(fc.df$cc))

total.metabs
```

### Theoretical distributions

`prob.sig` is the line/phase specific probability of being a metab with abs(l2fc) >= 1 given a total number of features (charge + compound).
```{r}
sig.probs <- fc.df %>% 
  mutate(direc = ifelse(l2ratio > 0, "Up", "Down")) %>% 
  filter(sig == TRUE) %>% 
  group_by(clean, phase, direc) %>% 
  tally() %>% 
  ungroup() %>% 
  mutate(prob.sig = n / total.metabs) %>% 
  group_by()
```

Get the theoretical distributions and use that to figure out how many features you'd expect to see altered in the same direction (`shared_metabs`) in `n_lines`.
```{r}
sinib.probs <- sig.probs %>% 
  split(list(.$phase, .$direc)) %>% # for each phase and direction
  map(function(x){
    # what is the prob of picking a metab that is sig in n lines
    dsinib(x = 1:12, # chance of getting 1, 2, ... successes, i.e. same metab same direc in n lines
           size = rep(1, 12), # pick one metab, 12 times, one for each of the 1:12 above
           prob = x$prob.sig) %>% # probs from the data
      as_tibble() %>% # convert to df
      mutate(n_lines = 1:12) # add n lines values to match the probs
  }) %>% 
  bind_rows(.id = "sample") %>% 
  separate(sample, into = c("phase", "direc"), sep = "\\.") %>% 
  mutate(type = "theoretical") %>% 
  rename("probs" = "value") %>% 
  mutate(shared_metabs = floor(total.metabs * probs)) # given these probs, how many metabs should be sig in n lines? 

sinib.probs
```

Get the actual number distributions my multiplying `n_lines` by the `shared_metabs` in a phase and direction specific manner.
```{r}
theory.dist <- sinib.probs %>% 
  split(list(.$phase, .$direc)) %>% 
  map(function(x){
    rep(
      x$n_lines,
      x$shared_metabs
    )
  })

theory.dist
```

### Observed

Observed probability of being shared, add a 1-sum here
```{r}
obs.probs <- fc.df %>%
  filter(sig == TRUE) %>%
  mutate(direc = ifelse(l2ratio > 0, "Up", "Down")) %>%
  group_by(cc, direc, phase) %>%
  tally() %>%
  ungroup() %>% 
  group_by(direc, phase, n) %>% 
  tally() %>% 
  ungroup() %>% 
  mutate(probs = nn / total.metabs,
         type = "Observed") %>% 
  select(direc, phase, n, probs, type) %>% 
  rename("n_lines" = "n")

obs.probs
```

```{r}
# make a matching df of theoretical probs
th.probs <- sinib.probs %>% 
  select(phase, direc, n_lines, probs) %>% 
  mutate(type = "Theoretical")

th.probs
```

How many features are actually shared? 
```{r}
actual.dist <- fc.df %>%
  filter(sig == TRUE) %>%
  mutate(direc = ifelse(l2ratio > 0, "Up", "Down")) %>%
  group_by(cc, direc, phase) %>%
  tally() %>%
  ungroup() %>% 
  split(list(.$phase, .$direc)) %>% 
  map(function(x){
    pull(x, n) %>% 
      sort()
  })

actual.dist
```

Run t-test to compare theoretical and observed
```{r}
t.res <- sapply(names(theory.dist), function(x) {
  t.test(theory.dist[[x]],
         actual.dist[[x]]) %>%
    tidy()
}, simplify = FALSE) %>%
  bind_rows(.id = "sample") %>% 
  separate(sample, into = c("phase", "direc"), sep = "\\.") %>% 
  mutate(xpos = 6.5, # these are for positioning them on a graph
         ypos = 60,
         p.value = paste("p =", signif(p.value, 1))) %>% 
  mutate(phase = ifelse(phase == "e", "Exponential", "Stationary"),
         type = "Observed")

t.res
```

Make a graph of the probabilities 
```{r fig.width = 6.51, fig.height = 4}
pdf <- bind_rows(th.probs, obs.probs) %>%
  mutate(phase = ifelse(phase == "e", "Exponential", "Stationary"))

p.lines <- pdf %>% 
  ggplot(., aes(n_lines, probs, color = direc, linetype = type)) +
  geom_line() +
  geom_point() +
  facet_grid(direc ~ phase) +
  scale_x_continuous(breaks = 0:12) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        text = element_text(size = 12),
        legend.position = c(.88, .1),
        legend.background = element_blank()) +
  scale_color_manual(
    values = c("indianred", "steelblue"),
    name = NULL,
    guide = "none"
  ) +
  labs(x = "Number of evolved lines",
       y = "Probability") +
  scale_linetype_discrete(name = NULL) +
  scale_y_log10(labels = trans_format("log10", math_format(10 ^ .x)))

p.lines
```

```{r}
ggsave(plot = p.lines, filename = "../../figures/probs_supp_1.pdf", width = 6.51, height = 4)
```

### Number of metabs altered

How many metabs did we expect to and actually see altered in the same direction?
```{r}
t.metabs <- sinib.probs %>% 
  select(phase, direc, n_lines, shared_metabs)

o.metabs <- fc.df %>%
  filter(sig == TRUE) %>%
  mutate(direc = ifelse(l2ratio > 0, "Up", "Down")) %>%
  group_by(cc, direc, phase) %>%
  tally(name = "n_lines") %>%
  ungroup() %>% 
  group_by(direc, phase, n_lines) %>% 
  tally(name = "shared_metabs") %>% 
  ungroup()
```

```{r}
plot.df <- bind_rows("Theoretical" = t.metabs, "Observed" = o.metabs, .id = "type") %>%
  mutate(phase = ifelse(phase == "e", "Exponential", "Stationary"))
```

As lines
```{r}
p.lines2 <- plot.df %>%
  ggplot(., aes(
    factor(n_lines),
    shared_metabs,
    group = type,
    color = direc,
    linetype = type
  )) +
  geom_point()+
  geom_line()+
  facet_grid(direc ~ phase) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    text = element_text(size = 12),
    legend.position = c(.89, .92),
    legend.background = element_blank()
  ) +
  scale_color_manual(
    values = c("indianred", "steelblue"),
    name = NULL,
    guide = "none"
  ) +
  scale_linetype_discrete(name = NULL)+
  labs(x = "Number of evolved lines",
       y = "Number of shared metabolic features")+
  geom_text(
    inherit.aes = FALSE,
    data = t.res,
    aes(
      x = xpos,
      y = ypos,
      label = p.value
    ),
    show.legend = FALSE
  )

p.lines2
```

```{r}
ggsave(plot = p.lines2, filename = "../../figures/probs_supp_2.pdf", width = 6.51, height = 4)
```

```{r}
sessionInfo()
```