---
title: "Standards and correlations"
output: 
  html_document:
    df_print: paged
    code_folding: hide
author: "John Favate"
date: "`r Sys.time()`"
---

<style type="text/css">
.main-container {
  max-width: 1500px;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r}
# Prevent printing of warnings and such in the HTML
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.align = "center")
```

```{r}
library(tidyverse)
library(patchwork)
library(scales)
library(ggpubr)
```

```{r}
df <- read_csv("../../data_frames/targeted_with_imps.csv")

# the untargeted data
untarg <- read_csv("../../data_frames/untargeted_data.csv")
```

This is the order I want the samples in, namely grouped by growth phase and replicates next to each other.
```{r}
sample.order <- names(df)[!(names(df) %in% c("compound", "charge", "is_standard"))] %>% 
  as_tibble() %>% 
  separate(value, into = c("line", "phase", "repl"), sep = "_") %>% 
  arrange(phase, line, repl) %>% 
  unite("sample", c(line, phase, repl)) %>% 
  pull(sample)
```

df for the labels
```{r}
lab.df <- data.frame(
  charge = "negative",
  x = c(16e-8, 11e-8, 11e-8),
  y = c(135, 115, 95),
  colz = c("Exponential", "Stationary", "Standards"),
  labz = c("Exponential", "Stationary", "Standards")
)
```

```{r}
dist.labels <- c("FALSE" = "metabolites", "TRUE" = "standards", "positive" = "positive", "negative" = "negative")

p.dists <- df %>% 
  select(-peak_area) %>% 
  pivot_longer(cols = where(is.numeric)) %>% 
  mutate(color = ifelse(is_standard == TRUE, "Standards",
                        ifelse(phase == "e", "Exponential", "Stationary")),
         color = factor(color, levels = c("Exponential", "Stationary", "Standards"))) %>% 
  ggplot(., aes(value, color = color))+
  facet_wrap(~ charge, scales = "free", labeller = as_labeller(dist.labels), ncol = 2)+
  geom_freqpoly(bins = 100)+
  scale_x_log10(labels = trans_format("log10", math_format(10^.x)))+
  labs(x = "Normalized peak area",
       y = "Count")+
  scale_color_manual(values = c("#B2DF8A", "#33A02C", "black"), name = NULL, guide = "none")+
  theme_bw()+
  theme(text = element_text(size = 12),
        panel.grid = element_blank(),
        legend.background = element_blank())+
  geom_text(inherit.aes = FALSE, data = lab.df, aes(x, y, color = colz, label = labz))

p.dists
```

### Correlations

Sample to sample correlations, except split by who being compared to who (i.e. anc to evo, ex to st, etc). What's compared is glucose_negative between samples, meaning compound+charge are the variables here
```{r}
cor.input <- df %>%
  filter(is_standard == FALSE) %>%
  unite("cc", charge, compound, sep = "_") %>%
  unite("sample", line, phase, repl, sep = "_") %>%
  select(cc, sample, n_peak_area) %>%
  mutate(n_peak_area = log10(n_peak_area)) %>%
  pivot_wider(names_from = sample, values_from = n_peak_area) %>%
  column_to_rownames("cc")

cor.res <- cor(cor.input)

# remove redundancy
cor.res[lower.tri(cor.res)] <- NA

# reshape to a tidy data frame with complete info on line, phase, repl, etc
cor.df <- as_tibble(cor.res, rownames = "samp1") %>%
  pivot_longer(where(is.numeric),
               names_to = "samp2",
               values_to = "R") %>%
  filter(samp1 != samp2 & !is.na(R)) %>%
  separate(samp1,
           into = c("line1", "phase1", "repl1"),
           sep = "_") %>%
  separate(samp2,
           into = c("line2", "phase2", "repl2"),
           sep = "_")

# this uses this list of strings which are eventually interpreted
# to create a list of sets of cors that are eventually joined
filter.vecs <- list(
  "Pair-\nwise" = "!is.na(R)",
  "Repli-\ncates" = "line1 == line2 & phase1 == phase2",
  "Intra-\nphase" = "phase1 == phase2",
  "Inter-\nphase" = "line1 == line2 & phase1 != phase2"
)

# lapply over this list to group the different kinds of correlations
marked.cors <- lapply(filter.vecs, function(y) {
  cor.df %>%
    filter(eval(parse(text = y)))
}) %>%
  bind_rows(.id = "type")

# reduce it so the df is smaller for saving as an rds
pa.df <- marked.cors %>% 
  select(type, R) %>% 
  mutate(type = factor(type, names(filter.vecs)))
```

Plot it.
```{r}
comps <- list(c("Intra-\nphase", "Inter-\nphase"))

p.cors <- pa.df %>%
  ggplot(., aes(type, R)) +
  geom_boxplot(outlier.size = 1, color = "black") +
  theme_bw() +
  theme(
    text = element_text(size = 13),
    legend.position = c(.85, .13),
    legend.background = element_blank(),
    panel.grid = element_blank()
  ) +
  labs(x = NULL) +
  scale_y_continuous(limits = c(.4, 1.1)) +
  stat_compare_means(
    comparisons = comps,
    vjust = -.5,
    method.args = list(alternative = "greater")
  )

p.cors
```

Sample to sample correlations, except split by who being compared to who (i.e. anc to evo, ex to st, etc). What's compared is glucose_negative between samples, meaning compound+charge are the variables here. This time, with means. 
```{r}
cor.input2 <- df %>%
  filter(is_standard == FALSE) %>%
  group_by(charge, line, phase, compound) %>%
  summarise(mean_pa = mean(n_peak_area)) %>%
  ungroup() %>%
  unite("cc", charge, compound, sep = "_") %>%
  unite("sample", line, phase, sep = "_") %>%
  select(cc, sample, mean_pa) %>%
  mutate(mean_pa = log10(mean_pa)) %>%
  pivot_wider(names_from = sample, values_from = mean_pa) %>%
  column_to_rownames("cc")

cor.res2 <- cor(cor.input2)

# remove redundancy
cor.res2[lower.tri(cor.res2)] <- NA

# reshape to a tidy data frame with complete info on line, phase, repl, etc
cor.df2 <- as_tibble(cor.res2, rownames = "samp1") %>%
  pivot_longer(where(is.numeric),
               names_to = "samp2",
               values_to = "R") %>%
  filter(samp1 != samp2 & !is.na(R)) %>%
  separate(samp1,
           into = c("line1", "phase1"),
           sep = "_") %>%
  separate(samp2,
           into = c("line2", "phase2"),
           sep = "_")

# this uses this list of strings which are eventually interpreted
# to create a list of sets of cors that are eventually joined
filter.vecs <- list(
  "Pair-\nwise" = "!is.na(R)",
  "Intra-\nphase" = "phase1 == phase2",
  "Inter-\nphase" = "line1 == line2 & phase1 != phase2"
)

# lapply over this list to group the different kinds of correlations
marked.cors2 <- lapply(filter.vecs, function(y) {
  cor.df2 %>%
    filter(eval(parse(text = y)))
}) %>%
  bind_rows(.id = "type")

pb.df <- marked.cors2 %>%
  select(type, R) %>% 
  mutate(type = factor(type, names(filter.vecs)))
```

Plot it.
```{r}
p.cors2 <- pb.df %>%
  ggplot(., aes(type, R)) +
  geom_boxplot(outlier.size = 1, color = "black") +
  theme_bw() +
  theme(
    text = element_text(size = 13),
    legend.position = c(.85, .13),
    legend.background = element_blank(),
    panel.grid = element_blank()
  ) +
  labs(x = NULL) +
  stat_compare_means(comparisons = comps,
                     vjust = -.5,
                     method.args = list(alternative = "greater")) +
  scale_y_continuous(limits = c(.4, 1.1))

p.cors2
```

### Comparison of values across ionization modes

get compounds that are in both +/- mode
```{r}
in.both <- df %>% 
  select(charge, compound) %>% 
  unique() %>% 
  count(compound) %>% 
  filter(n == 2) %>% 
  pull(compound)
```

Mean peak areas for the compounds
```{r}
mean.df <- df %>% 
  filter(is_standard == FALSE & compound %in% in.both) %>% 
  select(charge, line, repl, compound, n_peak_area) %>% 
  group_by(charge, compound, line) %>% 
  summarise(l10mpa = log10(mean(n_peak_area))) %>% 
  ungroup() %>% 
  pivot_wider(names_from = charge, values_from = l10mpa)

mean.df
```

They're definitely not 1:1 but they sometimes agree, sometimes they don't though. Acetyl-methionine doesn't agree across modes. Or cytosine.
```{r fig.width = 6.5, fig.height = 7}
mean.df %>% 
  ggplot(., aes(positive, negative))+
  geom_abline(aes(slope = 1, intercept = 0))+
  geom_point(size = 1)+
  facet_wrap(~line, scales = "free")+
  ggpubr::stat_cor(size = 3)+
  theme_bw()
```

```{r}
charge.cors <- mean.df %>%
  split(.$line) %>%
  map(function(x) {
    tmp <- x %>%
      select(positive, negative) %>%
      cor() %>%
      min(na.rm = TRUE)
    
    data.frame(R = tmp)
  }) %>%
  bind_rows(.id = "compound")

charge.cors
```

plot it
```{r}
p.charge.r <- charge.cors %>% 
  ggplot(., aes(x = "", R))+
  geom_boxplot(outlier.color = NA)+
  geom_jitter(height = 0, width = .2, size = 1)+
  theme_bw()+
  theme(text = element_text(size = 13),
        panel.grid = element_blank(),
        axis.ticks.x = element_blank())+
  labs(x = NULL,
       y = "R of ion modes")

p.charge.r
```


```{r fig.width = 6.51, fig.height = 5}
top <- p.dists

bottom <- (p.cors | p.cors2 | p.charge.r) + plot_layout(widths = c(.5, .4, .1))

ff <- (top / bottom) + plot_annotation(tag_levels = "A")

ff
```

```{r}
ggsave(plot = ff, filename = "../../figures/standards_and_cors.pdf", width = 6.51, height = 5)
```

### Untargeted data

How much stuff is in the untargeted data? Depends how you count, i.e. does it have a name, formula, MW, RT, all of them, some of them, etc.
```{r}
untarg
```

