---
title: "Standards and correlations"
output: 
  html_document:
    df_print: paged
    code_folding: hide
author: "John Favate"
date: "`r Sys.time()`"
---

<style type="text/css">
.main-container {
  max-width: 1500px;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r}
# Prevent printing of warnings and such in the HTML
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.align = "center")
```

```{r}
library(tidyverse)
library(patchwork)
library(scales)
library(ggpubr)
```

```{r}
df <- read_csv("../../data_frames/targeted_with_imps.csv") %>% 
  filter(is_standard == FALSE)

clean.names <- data.frame(line = unique(df$line)) %>%
  mutate(clean = case_when(
    grepl("ap", line) ~ str_replace(line, "ap", "A+"),
    grepl("am", line) ~ str_replace(line, "am", "A-"),
    grepl("rel", line) ~ str_replace(line, "rel60", "R0")
  ))

mut.lines <- c("A-1", "A-2", "A-3", "A-4", "A+3", "A+6")

df <- left_join(df, clean.names)
```

### Distribution of normalized peak areas

```{r fig.width = 6.51, fig.height = 5}
ann.df <- data.frame(
  phase = rep("Exponential", 2),
  x = c(1.2, 2.4),
  y = rep(5e-8, 2),
  label = c("R1", "R2")
)

p.dists <- df %>%
  mutate(
    charge = ifelse(charge == "positive", "(+)", "(-)"),
    phase = ifelse(phase == "e", "Exponential", "Stationary")
  ) %>%
  unite("samp", clean, charge, sep = " ") %>%
  ggplot(., aes(samp, n_peak_area, fill = repl)) +
  geom_boxplot(outlier.size = .2, show.legend = FALSE) +
  scale_y_log10(labels = trans_format("log10", math_format(10 ^ .x))) +
  facet_wrap( ~ phase, ncol = 1) +
  theme_bw() +
  theme(
    text = element_text(size = 12),
    axis.text.x = element_text(angle = 90, vjust = .5),
    panel.grid = element_blank()
  ) +
  labs(x = NULL,
       y = "Normalized peak area") +
  geom_text(
    inherit.aes = FALSE,
    data = ann.df,
    aes(
      x = x,
      y = y,
      label = label,
      color = label
    ),
    show.legend = FALSE
  )

p.dists
```


### Correlations

Sample to sample correlations, except split by who being compared to who (i.e. anc to evo, ex to st, etc). What's compared is glucose_negative between samples, meaning compound+charge are the variables here
```{r}
cor.input <- df %>%
  filter(is_standard == FALSE) %>%
  unite("cc", charge, compound, sep = "_") %>%
  unite("sample", line, phase, repl, sep = "_") %>%
  select(cc, sample, n_peak_area) %>%
  mutate(n_peak_area = log10(n_peak_area)) %>%
  pivot_wider(names_from = sample, values_from = n_peak_area) %>%
  column_to_rownames("cc")

cor.res <- cor(cor.input)

# remove redundancy
cor.res[lower.tri(cor.res)] <- NA

# reshape to a tidy data frame with complete info on line, phase, repl, etc
cor.df <- as_tibble(cor.res, rownames = "samp1") %>%
  pivot_longer(where(is.numeric),
               names_to = "samp2",
               values_to = "R") %>%
  filter(samp1 != samp2 & !is.na(R)) %>%
  separate(samp1,
           into = c("line1", "phase1", "repl1"),
           sep = "_") %>%
  separate(samp2,
           into = c("line2", "phase2", "repl2"),
           sep = "_")

# this uses this list of strings which are eventually interpreted
# to create a list of sets of cors that are eventually joined
filter.vecs <- list(
  "Pairwise" = "!is.na(R)",
  "Replicates" = "line1 == line2 & phase1 == phase2",
  "Intraphase" = "phase1 == phase2",
  "Interphase" = "line1 == line2 & phase1 != phase2"
)

# lapply over this list to group the different kinds of correlations
marked.cors <- lapply(filter.vecs, function(y) {
  cor.df %>%
    filter(eval(parse(text = y)))
}) %>%
  bind_rows(.id = "type")

# reduce it so the df is smaller for saving as an rds
pa.df <- marked.cors %>% 
  select(type, R) %>% 
  mutate(type = factor(type, names(filter.vecs)))
```

Plot it.
```{r}
comps <- list(c("Intraphase", "Interphase"))

p.cors <- pa.df %>%
  ggplot(., aes(type, R)) +
  geom_boxplot(outlier.size = .5, color = "black") +
  theme_bw() +
  theme(
    text = element_text(size = 13),
    legend.position = c(.85, .13),
    legend.background = element_blank(),
    panel.grid = element_blank()
  ) +
  labs(x = NULL) +
  scale_y_continuous(limits = c(0, 1.1),
                     breaks = seq(0, 1, .25),
                     name = "Pearson correlation of\nmetabolite abundances") +
  stat_compare_means(
    comparisons = comps,
    vjust = -.2,
    method.args = list(alternative = "greater")
  )

p.cors
```


### Comparison of values across ionization modes

get compounds that are in both +/- mode
```{r}
in.both <- df %>% 
  select(charge, compound) %>% 
  unique() %>% 
  count(compound) %>% 
  filter(n == 2) %>% 
  pull(compound)
```

Mean peak areas for the compounds
```{r}
mean.df <- df %>% 
  filter(is_standard == FALSE & compound %in% in.both) %>% 
  select(charge, line, repl, compound, n_peak_area) %>% 
  group_by(charge, compound, line) %>% 
  summarise(l10mpa = log10(mean(n_peak_area))) %>% 
  ungroup() %>% 
  pivot_wider(names_from = charge, values_from = l10mpa)

mean.df
```

They're definitely not 1:1 but they sometimes agree, sometimes they don't though. Acetyl-methionine doesn't agree across modes. Or cytosine.
```{r fig.width = 6.5, fig.height = 7}
mean.df %>% 
  ggplot(., aes(positive, negative))+
  geom_abline(aes(slope = 1, intercept = 0))+
  geom_point(size = 1)+
  facet_wrap(~line, scales = "free")+
  ggpubr::stat_cor(size = 3)+
  theme_bw()
```

```{r}
charge.cors <- mean.df %>%
  split(.$line) %>%
  map(function(x) {
    tmp <- x %>%
      select(positive, negative) %>%
      cor() %>%
      min(na.rm = TRUE)
    
    data.frame(R = tmp)
  }) %>%
  bind_rows(.id = "compound")

charge.cors
```

plot it
```{r}
p.charge.r <- charge.cors %>%
  ggplot(., aes(x = "", R)) +
  geom_boxplot(outlier.color = NA) +
  geom_jitter(height = 0,
              width = .2,
              size = 1) +
  theme_bw() +
  theme(
    text = element_text(size = 13),
    panel.grid = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  labs(x = NULL,
       y = "R of ion modes") +
  scale_y_continuous(limits = c(0, 1.1),
                     breaks = seq(0, 1, .25),
                     name = "Pearson correlation of\n ionization modes")

p.charge.r
```


```{r fig.width = 6.51, fig.height = 6.4}
bottom <- p.dists

top <- (p.cors | p.charge.r) + plot_layout(widths = c(.8, .2))

ff <- (top / bottom) + plot_annotation(tag_levels = "A") + plot_layout(heights = c(.44, .56))

ff
```

```{r}
ggsave(plot = ff, filename = "../../figures/standards_and_cors.pdf", width = 6.51, height = 6.4)
```